Analyzing Information Transfer from Heterogeneous Sources via Precise High-dimensional Asymptotics

### 2021-07-25 revision_journal/intro.tex
finish intro

%Many modern machine learning methods
%Hard parameter sharing (HPS) for multi-task learning is widely used in empirical research and goes back to the seminal work of \cite{C97}.
%Recent work has revived interests in this approach because it improves performance and reduces the cost of collecting labeled data \cite{R17}.
%It is generally applied by sharing the feature layers between all tasks while keeping an output layer for every task.
%Often, hard parameter sharing offers two critical advantages if successfully applied.
%First, it reduces model parameters since all tasks use the same feature space.
%Second, it reduces the amount of labeled data needed from each task by augmenting the entire training dataset.

%Hard parameter sharing works as an inductive transfer mechanism and a regularizer that reduces overfitting, both of which have great intuitive appeal \cite{R17}.
%For example, by restricting the shared space's size, HPS encourages information sharing among multiple tasks \cite{KD12}.
%Another source of inductive bias comes from the tasks and depends on datasets' properties such as sample sizes and task covariances \cite{WZR20}.
%However, how these dataset properties impact HPS has not been well established.
%It becomes increasingly important to understand HPS' formal generalization properties.
%Part of the challenge may be that HPS' generalization performance depends intricately on the sample size ratios and covariate shifts between tasks, and is not amenable to standard concentration results.
%Previous results based on Rademacher complexity or VC dimensions have considered cases where all tasks' sample sizes are equal to logarithmic factors of the feature dimension \cite{B00,MPR16}, and when all tasks' sample sizes increase simultaneously \cite{AZ05,M06}.
%For, the generalization error scales down as the sample sizes of all tasks increase, when applied to the multi-task setting \cite{B00,AZ05,M06,MPR16,WZR20}.

%This paper presents new techniques to study hard parameter sharing and establishes a number of new results.
%We consider regression analysis, which is arguably one of the most fundamental problems in statistics and machine learning.
%We are interested in the \textit{high-dimensional} setting, where each dataset's sample size and feature dimension grow linearly instead of logarithmically.
%This setting captures the fact that a single task's sample size is usually insufficient for accurate learning in many applications.
%For example, if a dataset's sample size is only a constant factor of dimension in linear regression, the variance is also constant (cf. Fact \ref{fact_tr}).
%The high-dimensional setting is challenging but is crucial for understanding how datasets' sample sizes impact generalization performance.

@@

We first define the data model.
Suppose we have two datasets (or tasks).
We refer the first dataset as the source and the second task as the target.
Let $i = 1$ or $2$.
Let $n_i$ be dataset $i$'s sample size.
Let $x_1^{(i)}, x_2^{\ss (i)}, \dots, x_{n_i}^{(i)}$ be the $n_i$ datapoints' $p$-dimensional covariates.
Let $y_1^{(i)}, y_2^{(i)}, \dots, y_{n_i}^{(i)}$ be these datapoints' labels.
We assume a linear model specified by an unknown model vector $\beta^{(i)} \in \real^p$ as follows:
\begin{align}
    y_j^{(i)} = \inner{x_j^{(i)}}{\beta^{(i)}} + \varepsilon_j^{(i)}, \text{ for any } j = 1,\dots,n_i, \label{eq_linear}
\end{align}
where $\varepsilon^{(i)}_j\in\real$ denotes a random noise variable with mean zero and variance $\sigma^2$.
Our goal is to learn an estimator to predict the target task given both datasets.

We learn a two-layer linear neural network with parameters $A \in \real^{p\times r}$, $B_1 \in \real^{r}$, and $B_2 \in \real^r$ by minimizing the following optimization objective:
\begin{align}\label{eq_hps}
    f(A, B_1, B_2) = \bignorm{X^{(1)} A B_1 - Y^{(1)}}_2^2 + \bignorm{X^{(2)} A B_2 - Y^{(2)}}_2^2.
\end{align}
The above two-layer neural network uses a shared feature space $A$ for both tasks while using a separate prediction head for each task.\footnote{Popular variants of the objective function \eqref{eq_hps} include adding a weight parameter to each summand and adding ridge regularization. We focus on the unweighted and unregularized objective for simplicity. All of the asymptotic limits we show below can be straightforwardly extended to weighted and regularized objectives.}
Such models are also known as \textit{hard parameter sharing} in the machine learning literature \cite{C97,R17}.
Suppose that $(\hat{A}, \hat{B}_1, \hat{B}_2)$ is a global minimizer of $f(\cdot)$.
We define the hard parameter sharing (HPS) estimator for the target task as $\hat{\beta}_2^{\MTL} \define \hat{B} \hat{A}_2$.
In order to evaluate the estimator, we will study the excess risk at a new (unseen) test datapoint from task two, denoted as $L(\hat{\beta}_2^{\MTL})$ (see Section \ref{sec_risk} for its precise definition).

As a motivating example of the above setup, we show an intriguing simulation where combining both datasets jointly results in different transfer effect on $L(\hat{\beta}_2^{\MTL})$, depending on the distance between $\beta^{(1)}$ and $\beta^{(2)}$.
Figure \ref{fig_motivation} shows the result for a setting where every entry of $X_1, X_2$ is generated from an isotropic Gaussian, and $\beta_1,\beta_2$ are generated such that $\norm{\beta^{(1)} - \beta^{(2)}}^2 \approx 2\mu^2$.
We observe that for $\mu = 0.25$, $L(\hat{\beta}_2^{\MTL})$ is always smaller than the excess risk of OLS.
For $\mu = 0.3$ or $0.35$, $L(\hat{\beta}_2^{\MTL})$ is smaller than the excess risk of OLS for a restricted range of $n_1$.
Such a scenario, where combining both datasets jointly improves over learning the target task individually, is also called a \textit{positive information transfer} \cite{PY09}.
On the other hand, for $\mu = 0.45$, HPS is always worse than OLS, thus providing a \textit{negative information transfer}.  
Our main contribution is to precisely analyze a number of interesting phenomena including the above that are prevalent in transfer learning, depending on geometries of the covariance matrices and model distances.

\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/motivation.eps}
    \caption{We illustrate that our setup can reproduce the phenomenon of \textit{positive} (or \textit{negative}) \textit{information transfer}, which is prevalent in the context of transfer learning.
    We vary the sample size of dataset one and the distance parameter $\mu$ such that $\norm{\beta^{(1)} - \beta^{(2)}}^2 \approx 2\mu^2$. The region below the excess risk of the ordinary least squares (OLS) estimator corresponds to \textit{positive information transfer} when $L(\hat{\beta}_2^{\MTL})$ is smaller than the excess risk of the OLS estimator. This simulation uses $p = 100, n_2 = 300, \sigma = 1/2$. See also Figure \ref{fig_ab_data} for a similar phenomenon observed in text classification tasks.}
    \label{fig_motivation}
\end{figure*}

@@

We consider a high-dimensional setting on each dataset where the sample size $n_i$ increases with $p$ to infinity proportionally by a constant factor, which has recently received lots of interests (see e.g., \citet{hastie2019surprises} and Section \ref{sec_related} for more references).
Let $\Sigma^{(i)} \in \real^{p\times p}$ be a deterministic positive semidefinite matrix.
The covariates are assumed to have population covariance equal to $\Sigma^{(i)}$:
\begin{align}
    x_j^{(i)} = \big(\Sigma^{(i)}\big)^{1/2} z_{j}^{(i)}, \text{ for any } j = 1,\dots,n_i, \label{eq_rvm}
\end{align}
where $z_j^{(i)}$ consists of independent and identical entries sampled from a distribution that has zero mean, unit variance, and satisfies certain bounded moment condition (see Section \ref{sec_data} for the precise assumptions).

We focus on the so called \textit{underparametrized} setting for the target task where $n_2 > p$ whereas for dataset one $n_1$ can be either smaller or greater than $p$.
Thus, the OLS estimator exists and it is well-known that the limit of its excess risk when $p$ approaches infinity is equal to $\frac{\sigma^2 p}{n_2 - p}$ (see e.g., \citet{bai2009spectral}).
By contrast, showing the asymptotic limit of $L(\hat{\beta}^{\MTL}_2)$ in the high-dimensional setting requires dealing with the inverse of the sum of two sample covariance matrices.
The first challenge is dealing with \textit{covariate shift}---when $\Sigma^{(1)}$ and $\Sigma^{(2)}$ are different.
The second challenge is dealing with \textit{model shift}---when $\beta^{(1)}$ and $\beta^{(2)}$ are different.
Our main result addresses both challenges by \todo{illustrate}.
\begin{itemize}
    \item \textbf{Covariate shift:}
    Our first result applies to settings where $\Sigma^{(1)}$ and $\Sigma^{(2)}$ can be arbitrarily different but $\beta^{(1)} = \beta^{(2)}$.
    For such settings, we show in Theorem \ref{thm_main_RMT} the asymptotic limit of $L(\hat{\beta}_2^{\MTL})$ as a function of the singular values of the matrix $\big(\Sigma^{(1)}\big)^{1/2}\big(\Sigma^{(2)}\big)^{-1/2}$ and the sample sizes $n_1, n_2$.
    \item \textbf{Model shift:} Our second result applies to settings where $\beta^{(1)}$ and $\beta^{(2)}$ can be arbitrarily different but $\Sigma^{(1)} = \Sigma^{(2)}$.
    For such settings, we show in Theorem \ref{cor_MTL_loss} the asymptotic limit of $L(\hat{\beta}_2^{\MTL})$ as a function of the $\ell_2$ distance between $\beta^{(1)}$ and $\beta^{(2)}$, and the sample sizes.
    \item \textbf{Covariate and model shift:} Our third result applies to settings with both covariate shift and model shift.
    When $\Sigma^{(2)}$ is isotropic, we show in Theorem \ref{???} the asymptotic limit of $L(\hat{\beta}_2^{\MTL})$ as a function of the singular values of $\Sigma^{(1)}$, the $\ell_2$ distance between $\beta^{(1)}$ and $\beta^{(2)}$), and the sample sizes.
    For the most general setting when $\Sigma^{(2)}$ is unisotropic, we show in Theorem \ref{prop_main_RMT} an estimate of $L(\hat{\beta}_2^{\MTL})$ that is accurate when $n_1$ is much larger than $n_2$.
\end{itemize}

Next, we demonstrate the power of these precise asymptotics by explaining a number of empirical phenomena in the context of transfer learning.
We consider a random-effect model where $\beta^{(i)}$ is equal to a shared model vector plus per-task variances (cf. \citet{dobriban2018high}).
Under this model, we show the following \textit{theoretical insights}.
\begin{enumerate}
    \item[i)] \textit{Whether or not covariate shift helps depends on sample sizes:} We consider the covariate shift setting and show that (cf. Proposition \ref{claim_dichotomy}) when $n_1 < n_2$, transferring from a data source with $\Sigma^{(1)} \neq \Sigma^{(2)}$ can achieve a smaller $L(\hat{\beta}_2^{\MTL})$ compared to transferring from a data source with $\Sigma^{(1)} = \Sigma^{(2)}$. When $n_1 > n_2$, transferring from from a data source with $\Sigma^{(1)} \neq \Sigma^{(2)}$ always results in higher $L(\hat{\beta}_2^{\MTL})$ than transferring from a data source with $\Sigma^{(1)} = \Sigma^{(2)}$.
    \item[ii)] \textit{Identical covariance provides approximately optimal transfer under imbalanced sizes:} Additionally, in Proposition \ref{prop_covariate}, we show that when $n_1$ is greater than $n_2$ times a certain large constant factor, transferring from a data source with $\Sigma^{(1)} = \Sigma^{(2)}$ achieves an approximately minimum excess risk value, within a certain family of covariance matrices of $\Sigma^{(1)}$.
    \item[iii)] \textit{Three regimes of information transfer under model shift}: Next, we consider the model shift setting and characterize the precise range of $n_1$ under which there is a positive transfer.
    In Proposition \ref{claim_model_shift}, we show that when $\mu^2 \le \frac{\sigma^2 p}{2(n_2 - p)}$ (recall that $\norm{\beta^{(1)} - \beta^{(2)}}^2 \approx 2\mu^2$) is small, the transfer effect is positive for any $n_1$.
    When $\frac{\sigma^2 p}{2(n_2 - p)} < \mu^2 < \frac{\sigma^2 n_2}{2(n_2 - p)}$, there is a transition from positive to negative transfer, as $n_1$ increases.
    When $\frac{\sigma^2 n_2}{2(n_2 - p)} \le \mu^2$, the transfer effect is negative for any $n_1$.
    Additionally, we illustrate that these three regimes of information transfer arises under both covariate and model shift (cf. Figure \ref{fig_sec3_cov_mo_b}).
\end{enumerate}

\paragraph{Empirical results.}
We also complement our theoretical analysis with empirical evaluations.
As mentioned in Section \ref{sec_mot}, our motivation for studying the HPS estimator stems from its superior empirical performance for transfer learning.
We affirm that HPS indeed outperforms several commonly used estimators including weighted averages of the OLS estimators of each task (cf. \citet{bastani2020predicting}) and the ridge estimator of the target task in our setting. See Figure \ref{fig_sec51} for the result.

@@

\paragraph{Extension to multiple sources.}
Finally, we also extend our setup to the case of having multiple data sources.
We consider a setting where all datasets have the same features but different labels, which is also known as multi-label prediction \cite{hsu2009multi}.
We show the asymptotic limit of $L(\hat{\beta}_i^{\MTL})$ under model shift (cf. Theorem \ref{thm_many_tasks}) and use the result to illustrate how to select the hidden width $r$ of $B$.
%Our result implies that when all tasks have the same features but different labels, for any task, HPS helps reduce the task's variance compared to single-task learning but increases bias.

@@

%Linear models in multi-task learning have been studied in various settings, including online learning \cite{CCG10,DCSP18}, sparse regression \cite{LPTV09,LPVT11}, and representation learning \cite{BHKL19}.

\paragraph{Random matrix theory in machine learning.}
This work is inspired by an emerging line of recent work that has found random matrix theory useful for studying modern machine learning techniques.
A prominent line of recent work in this direction has been to understand the generalization properties of interpolators \cite{bartlett2020benign,hastie2019surprises,montanari2019generalization,liang2020just,liang2020precise}.
Interpolators such as minimum-norm estimators (in the context of linear regression) share similar behavior with modern neural networks in the sense that such models learn by ``memorizing'' the data with lots of parameters \cite{mei2019generalization}.
The study of such interpolators has also led to an explanation of the double-descent phenomenon \cite{belkin2019reconciling}.

\paragraph{Sample covariance matrices and distribution shift.}
These previous works primarily deal with a single high-dimensional sample covariance matrix. %need to check 
In the \textit{underparametrized} setting ($n_2 > p$) which this work focuses on, when the covariates are sampled from Gaussian distributions, the precise asymptotic limit can be derived directly from properties of the Wishart distribution (see also Lemma \ref{fact_tr} and the historical notes in Section \ref{sec_bv}).
In the high-dimensional setting, the eigenvalues of Wishart matrices satisfy the well-known Marchenko–Pastur (MP) law \cite{MP}, whose Stieltjes transform characterizes the variance limit.
Furthermore, it is well-known that the MP law holds universally regardless of the underlying data distribution of the covariates \cite{bai2010spectral}. \citet{isotropic} obtained a sharp convergence rate of the empirical spectral distribution (ESD) of the MP law for sample covariance matrices with isotropic population covariance matrices. \citet{Anisotropic,DY} later extended this result to sample covariance matrices with arbitrary population covariances. These results are proved by establishing certain optimal convergence estimates of the Stieltjes transforms of sample covariance matrices, which are referred to as \emph{local laws} in the random matrix theory literature. We refer the reader to \cite{erdos2017dynamical} and the references therein for a detailed review of related concepts.
Our technical contribution is to extend these techniques to the two-task setting and prove a local law for the sum of two sample covariance matrices with arbitrary covariate shifts.
Using the local law, we can then derive the variance limit with a precise dependence on the singular values of the covariate shift matrix $\big(\Sigma^{(1)}\big)^{1/2} \big(\Sigma^{(2)}\big)^{-1/2}$.

@@

\paragraph{Transfer learning and domain adaptation theory.}
Earlier works on transfer learning and domain adaptation use uniform convergence to obtain generalization bounds \cite{B00,BS03,M06}.
The influential paper by \citet{BBCK10} proposes a rigorous model for domain adaptation in classification (see also \citet{crammer2008learning} for an earlier result).
The main result provides a generalization bound for combining multiple data sources using an objective that averages over the risk of every datapoint.
This bound provides a principled way to reweight each data source by minimizing the bound.
From a technical perspective, uniform convergence based techniques provide upper bounds on the excess risks.
On the other hand, analyzing the empirical phenomena of information transfer such as the ones in Figure \ref{fig_motivation} requires tight lower bounds on the excess risks.

%One limitation of uniform convergence based techniques is that the results often assume that all  tasks have the same sample size, see e.g. \cite{B00,MPR16}.
%\citet{pontil2013excess}
%Moreover, these techniques do not apply to the high-dimensional setting because the results usually require a sample size of at least $p \log p$.
Finally, we discuss several concurrent works that study transfer learning in linear regression settings.
The work of \citet{li2020transfer} considers the problem of how to select useful data sources for transfer learning under model shift.
Their work provides an adaptive procedure that applies even when the set of useful data sources are unknown.
Our work differs from theirs in two aspects.
First, we considered a high-dimensional setting where the sample size increases with dimension proportionally.
Such a setting allows us to derive precise asymptotics of the excess risks, which is then used to explain a number of interesting phenomena in transfer learning.
Second, our result applies to covariate-shifted (and not necessarily Gaussian) source datasets whereas the setting of \citet{li2020transfer} assumes the covariates of all data sources and the target are drawn from a Gaussian distribution with the same population covariance matrix.
The work of \citet{lei2021nearoptimal} considers a setting of linear regression under distribution shift including covariate shift and model shift similar to our setting.
The difference is that our setting deals with a labeled target dataset whereas the result of \citet{lei2021nearoptimal} applies to settings with an unlabeled target dataset.
%One closely related work is \cite{WZR20}, which studied hard parameter sharing for two linear regression tasks.
%	\item \citet{kalan2020minimax}
%	\item \citet{lounici2011oracle}

>>>>

Given data from two related tasks, how much does combining both data for learning a joint model help predict one of the tasks of interest?
For example, suppose you have $n_1$ datapoints with $p$-dimensional covariates (sampled from a distribution $D_1$) and real-valued labels in the first task, and $n_2$ datapoints with $p$-dimensional covariates (sampled from a different distribution $D_2$) and real-valued labels in the second task. 
Does combining the $n_1 + n_2$ datapoints help learn a model whose performance is better than a model learned using the $n_2$ datapoints alone?
Such questions are prevalent in transfer learning, where one often uses data from related tasks to expand the sample size of a task of interest.
A frequent impediment to applying transfer learning is \textit{negative information transfer} if combining both data performs worse than learning a single task clone \cite{PY09}.
\textit{Positive information transfer} refers to the desirable scenario where a task of interest benefits when combined with another task.

The answer to the two questions above hinges on the tasks' heterogeneity.
For example, negative information transfer can happen if the label distribution of the second task conditional on the covariates deviates significantly from that of the first task.
Additionally, the ``information transfer effect'' also hinges on the covariates' distributions, $D_1$ and $D_2$.
Following the existing literature (e.g., \citet{kouw2018introduction}), we refer to settings where $D_1 \neq D_2$ as \textit{covariate shift} and settings where label distributions differ (conditional on the covariates) as \textit{model shift}.
Both covariate and model shifts are prevalent in modern datasets (see e.g., \citet{koh2021wilds}).

This paper studies a linear regression setting involving two tasks with covariate and model shifts.
We use a two-layer linear neural network to combine the data from both tasks and show that this network reproduces several interesting phenomena in transfer learning.
We analyze the prediction risk of this network and its transfer effect, providing precise conditions based on geometries of $D_1, D_2$, and the sample sizes $n_1, n_2$.
The analysis has also inspired several insights for understanding and mitigating covariate and model shifts.

The key ingredient of our analysis is showing the asymptotic limit of certain combinations of sample covariance matrices under distribution shift, using the local laws developed recently by \citet{isotropic} and \citet{BES_free1}.
After formally defining the specific problem that we tackle, in Section \ref{sec_related}, we provide a technical discussion of the related works in random matrix theory.

%Our main contribution is to precisely analyze a number of interesting phenomena including the above that are prevalent in transfer learning, depending on .
 %an analysis of the precise causes of negative transfer is of significant practical interest.

@@

Let $i = 1$ or $2$.
Let $x_1^{(i)}, x_2^{\ss (i)}, \dots, x_{n_i}^{(i)}$ be the covariates.
Let $y_1^{(i)}, y_2^{(i)}, \dots, y_{n_i}^{(i)}$ be the corresponding labels.
We assume a linear model specified by an unknown model vector $\beta^{(i)} \in \real^p$ as follows:
\begin{align}
    y_j^{(i)} = \inner{x_j^{(i)}}{\beta^{(i)}} + \varepsilon_j^{(i)}, \text{ for any } j = 1,\dots,n_i, \label{eq_linear}
\end{align}
where $\varepsilon^{(i)}_j\in\real$ denotes a random noise variable with mean zero and variance $\sigma^2$.
For the ease of presentation, we refer to the first dataset as the source and the second task as the target.
%Our goal is to learn an estimator to predict the target task given both datasets.

We learn a two-layer linear neural network with parameters $A \in \real^{p\times r}$, $B_1 \in \real^{r}$, and $B_2 \in \real^r$ by minimizing the following optimization objective:
\begin{align}\label{eq_hps}
    f(A, B_1, B_2) = \bignorm{X^{(1)} A B_1 - Y^{(1)}}_2^2 + \bignorm{X^{(2)} A B_2 - Y^{(2)}}_2^2.
\end{align}
The above objective uses $A$ as the shared feature space for both tasks and $B_1, B_2$ as the prediction head for each task.\footnote{Popular variants of the objective function \eqref{eq_hps} include adding a weight parameter to each summand and adding ridge regularization. We focus on the unweighted and unregularized objective for simplicity. All of the asymptotic limits we show below can be straightforwardly extended to weighted and regularized objectives.}
Such models are also known as \textit{hard parameter sharing} \cite{C97,R17}.
We focus on the case of $r = 1$---otherwise, the global minimizer of $f(A, B_1, B_2)$ reduces to learning each task alone, resulting in no transfer effect (cf. Proposition \ref{prop_large_r} in Section \ref{sec_risk}).
In this case, let $(\hat{A}, \hat{B}_1, \hat{B}_2)$ be a global minimizer of $f(\cdot)$.
The hard parameter sharing (HPS) estimator for the target task is defined as $\hat{\beta}_2^{\MTL} \define \hat{B} \hat{A}_2$.
In order to evaluate the estimator, we will study the excess risk, which is the prediction risk at an unseen datapoint (of task two) minus $\sigma^2$, denoted as $L(\hat{\beta}_2^{\MTL})$.

As a motivating example of the above setup, we show an intriguing simulation where combining both datasets results in different transfer effects.
Figure \ref{fig_motivation} illustrates a setting where $X_1, X_2$ are sampled from an isotropic Gaussian and $\beta_1,\beta_2$ are generated so that $\norm{\beta^{(1)} - \beta^{(2)}}^2 \approx 2\mu^2$.
We observe that for $\mu = 0.25$, the transfer effect is always positive---$L(\hat{\beta}_2^{\MTL})$ is always smaller than the excess risk of OLS.
For $\mu = 0.3$ or $0.35$, the transfer effect is positive only for a restricted range of $n_1$.
On the other hand, for $\mu = 0.45$, the transfer effect is always negative.  
Our result below will imply the exact conditions for positive information transfer, depending on $\mu$ and $n_1$.

\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/motivation.eps}
    \caption{We illustrate that our setup can reproduce the phenomenon of \textit{positive} (or \textit{negative}) \textit{information transfer}, which is prevalent in transfer learning.
    We vary the sample size of dataset one and the distance parameter $\mu$ such that $\norm{\beta^{(1)} - \beta^{(2)}}^2 \approx 2\mu^2$. The region below the excess risk of the ordinary least squares (OLS) estimator corresponds to \textit{positive information transfer}. This simulation uses $p = 100, n_2 = 300, \sigma = 1/2$. See also Figure \ref{fig_ab_data} for a similar phenomenon observed in text classification tasks.}
    \label{fig_motivation}
\end{figure*}
% when $L(\hat{\beta}_2^{\MTL})$ is smaller than the excess risk of the OLS estimator

@@

We consider a high-dimensional setting on each dataset where the sample size $n_i$ increases with $p$ to infinity proportionally by a constant factor, which has recently received lots of interests (see e.g., \citet{hastie2019surprises} and Section \ref{sec_related} for more references).
Let $\Sigma^{(i)} \in \real^{p\times p}$ be a deterministic positive semidefinite matrix.
The covariates are assumed to have population covariance equal to $\Sigma^{(i)}$: $x_j^{(i)} = \big(\Sigma^{(i)}\big)^{1/2} z_{j}^{(i)},$ for any $j = 1,\dots,n_i,$
where $z_j^{(i)}$ consists of independent and identical entries sampled from a distribution that has zero mean, unit variance, and satisfies certain bounded moment condition (see Section \ref{sec_data} for the precise assumptions).

\paragraph{Main results.} We focus on the so called \textit{underparametrized} setting for the target task where $n_2 > p$ whereas for dataset one $n_1$ can be either smaller or greater than $p$.
Thus, the OLS estimator exists and it is well-known that the limit of its excess risk when $p$ approaches infinity is equal to $\frac{\sigma^2 p}{n_2 - p}$ (see e.g., \citet{bai2009spectral}).
By contrast, showing the asymptotic limit of $L(\hat{\beta}^{\MTL}_2)$ in the high-dimensional setting requires dealing with the inverse of the sum of two sample covariance matrices under distribution shift.
%The first challenge is dealing with \textit{covariate shift}---when $\Sigma^{(1)}$ and $\Sigma^{(2)}$ are different.
%The second challenge is dealing with \textit{model shift}---when $\beta^{(1)}$ and $\beta^{(2)}$ are different.
Our main result addresses both challenges by \todo{illustrate}.
\begin{itemize}
    \item %\textbf{Covariate shift:}
    Our first result applies to settings where $\Sigma^{(1)}$ and $\Sigma^{(2)}$ can be arbitrarily different but $\beta^{(1)} = \beta^{(2)}$.
    In Theorem \ref{thm_main_RMT}, we characterize the asymptotic limit of $L(\hat{\beta}_2^{\MTL})$ as a function of the singular values of the matrix $\big(\Sigma^{(1)}\big)^{1/2}\big(\Sigma^{(2)}\big)^{-1/2}$ and the sample sizes $n_1, n_2$.
    \item %\textbf{Model shift:}
    Our second result applies to settings where $\beta^{(1)}$ and $\beta^{(2)}$ can be arbitrarily different but $\Sigma^{(1)} = \Sigma^{(2)}$.
    In Theorem \ref{cor_MTL_loss}, we provide the asymptotic limit of $L(\hat{\beta}_2^{\MTL})$ as a function of the $\ell_2$ distance between $\beta^{(1)}$ and $\beta^{(2)}$, and the sample sizes $n_1, n_2$.
    \item %\textbf{Covariate and model shift:}
    Our third result applies to settings where both the covariance matrices and the model vectors are different between the two tasks.
    In Theorem \ref{???}, we show that when $\Sigma^{(2)}$ is isotropic, the asymptotic limit of $L(\hat{\beta}_2^{\MTL})$ is characterized by the singular values of $\Sigma^{(1)}$, the $\ell_2$ distance between $\beta^{(1)}$ and $\beta^{(2)}$), and the sample sizes $n_1, n_2$.
    Next, in Theorem \ref{prop_main_RMT}, we consider the most general setting concerning an arbitrary $\Sigma^{(2)}$. We identify an estimate of $L(\hat{\beta}_2^{\MTL})$ that becomes more and more accurate as $n_1 / n_2$ increases.
\end{itemize}

\noindent\textit{Extension to multiple sources.}
We also extend our setup to the case of having multiple data sources.
We consider a setting where all datasets have the same features but different labels, which is also known as multi-label prediction \cite{hsu2009multi}.
We show the asymptotic limit of $L(\hat{\beta}_i^{\MTL})$ under model shift (cf. Theorem \ref{thm_many_tasks}) and use the result to illustrate how to select the hidden width $r$ of $B$.

\paragraph{Theoretical insights and empirical implications.} Next, we demonstrate the power of these precise asymptotics by explaining several empirical phenomena in transfer learning.
We consider a random-effect model where $\beta^{(i)}$ is equal to a shared model vector plus per-task variances (cf. \citet{dobriban2018high}).
Under this model, we show the following \textit{theoretical insights}.
\begin{enumerate}
    \item[i)] \textit{Whether or not covariate shift helps depends on sample sizes:} We consider the covariate shift setting and show that (cf. Proposition \ref{claim_dichotomy}) when $n_1 < n_2$, transferring from a data source with $\Sigma^{(1)} \neq \Sigma^{(2)}$ can achieve a smaller $L(\hat{\beta}_2^{\MTL})$ compared to transferring from a data source with $\Sigma^{(1)} = \Sigma^{(2)}$. When $n_1 > n_2$, transferring from from a data source with $\Sigma^{(1)} \neq \Sigma^{(2)}$ always results in higher $L(\hat{\beta}_2^{\MTL})$ than transferring from a data source with $\Sigma^{(1)} = \Sigma^{(2)}$.
    \item[ii)] \textit{Identical covariance provides approximately optimal transfer under imbalanced sizes:} Additionally, in Proposition \ref{prop_covariate}, we show that when $n_1$ is greater than $n_2$ times a certain large constant factor, transferring from a data source with $\Sigma^{(1)} = \Sigma^{(2)}$ achieves an approximately minimum excess risk value, within a certain family of covariance matrices of $\Sigma^{(1)}$.
    \item[iii)] \textit{Three regimes of information transfer under model shift}: Next, we consider the model shift setting and characterize the precise range of $n_1$ under which there is a positive transfer.
    In Proposition \ref{claim_model_shift}, we show that when $\mu^2 \le \frac{\sigma^2 p}{2(n_2 - p)}$ (recall that $\norm{\beta^{(1)} - \beta^{(2)}}^2 \approx 2\mu^2$) is small, the transfer effect is positive for any $n_1$.
    When $\frac{\sigma^2 p}{2(n_2 - p)} < \mu^2 < \frac{\sigma^2 n_2}{2(n_2 - p)}$, there is a transition from positive to negative transfer, as $n_1$ increases.
    When $\frac{\sigma^2 n_2}{2(n_2 - p)} \le \mu^2$, the transfer effect is negative for any $n_1$.
    Additionally, we illustrate that these three regimes of information transfer arises under both covariate and model shift (cf. Figure \ref{fig_sec3_cov_mo_b}).
\end{enumerate}

We also complement our theoretical analysis with empirical evaluations.
As mentioned in Section \ref{sec_mot}, our motivation for studying the HPS estimator stems from its superior empirical performance for transfer learning.
We affirm that HPS indeed outperforms several commonly used estimators including weighted averages of the OLS estimators of each task (cf. \citet{bastani2020predicting}) and the ridge estimator of the target task in our setting. See Figure \ref{fig_sec51} for the result.

@@

%Our result implies that when all tasks have the same features but different labels, for any task, HPS helps reduce the task's variance compared to single-task learning but increases bias.

@@

\paragraph{Sample covariance matrices under distribution shift.}
In the \textit{underparametrized} setting ($n_2 > p$) which this work focuses on, when the covariates are sampled from Gaussian distributions, the precise asymptotic limit can be derived directly from properties of the Wishart distribution (see also Lemma \ref{fact_tr} and the historical notes in Section \ref{sec_bv}).
In the high-dimensional setting, the eigenvalues of Wishart matrices satisfy the well-known Marchenko–Pastur (MP) law \cite{MP}, whose Stieltjes transform characterizes the variance limit.
Furthermore, it is well-known that the MP law holds universally regardless of the underlying data distribution of the covariates \cite{bai2010spectral}. \citet{isotropic} obtained a sharp convergence rate of the empirical spectral distribution (ESD) of the MP law for sample covariance matrices with isotropic population covariance matrices. \citet{Anisotropic,DY} later extended this result to sample covariance matrices with arbitrary population covariances. These results are proved by establishing certain optimal convergence estimates of the Stieltjes transforms of sample covariance matrices, which are referred to as \emph{local laws} in the random matrix theory literature. We refer interested readers to \citet{erdos2017dynamical} and the references therein for a detailed review of related concepts.
Our technical contribution is to extend these techniques to the two-task setting and prove a local law for the sum of two sample covariance matrices with arbitrary covariate shifts.
Using the local law, we can then derive the variance limit with a precise dependence on the singular values of the covariate shift matrix $\big(\Sigma^{(1)}\big)^{1/2} \big(\Sigma^{(2)}\big)^{-1/2}$.

@@

\paragraph{Transfer learning theory.}
Earlier works such as \citet{B00,BS03,M06} use uniform convergence to obtain generalization bounds on transfer learning (also called domain adaptation).
The influential paper by \citet{BBCK10} proposes a rigorous model for domain adaptation in classification (see also \citet{crammer2008learning} for an earlier result).
Their main result provides a generalization bound for combining multiple data sources using an objective that averages over the risk of every datapoint.
This bound provides a principled way to reweight each data source by minimizing the bound.
From a technical perspective, uniform convergence-based techniques provide upper bounds on the excess risks.
However, analyzing the empirical phenomena of information transfer such as the ones in Figure \ref{fig_motivation} requires tight lower bounds on the excess risks.
The very recent work of \citet{mousavi2020minimax} introduced a minimax framework to quantify the fundamental limit of transfer learning.
Their result quantities the minimax risk using the distance between the source and target models.
Our work provides precise estimates that accurately match the empirical risk of commonly used estimators compared to their work.
These precise estimates are necessary for analyzing these estimators' transfer effect.

%One limitation of uniform convergence based techniques is that the results often assume that all  tasks have the same sample size, see e.g. \cite{B00,MPR16}.
%\citet{pontil2013excess}
%Moreover, these techniques do not apply to the high-dimensional setting because the results usually require a sample size of at least $p \log p$.
Finally, we discuss several concurrent works that study transfer learning in linear regression settings.
The work of \citet{li2020transfer} considers the problem of how to select useful data sources for transfer learning under model shift.
Their work provides an adaptive procedure that applies even when the set of beneficial data sources is unknown.
Our work differs from theirs in two aspects.
First, we considered a high-dimensional setting where the sample size increases with dimension proportionally by a constant factor.
We use this setting to reproduce several interesting phenomena in  transfer learning.
Second, our estimates apply even under covariate shift, whereas the result of \citet{li2020transfer} requires that there is no covariate shift between the data sources and the target task.
Another work by \citet{lei2021near} considers a  linear regression setting under distribution shift, including covariate shift and model shift.
The difference is that our setting deals with a labeled target dataset. In contrast, \citet{lei2021near} deal with an unlabeled target dataset (i.e., unsupervised domain adaptation).
%Linear models in multi-task learning have been studied in various settings, including online learning \cite{CCG10,DCSP18}, sparse regression \cite{LPTV09,LPVT11}, and representation learning \cite{BHKL19}.

%One closely related work is \cite{WZR20}, which studied hard parameter sharing for two linear regression tasks.
%	\item \citet{kalan2020minimax}
%	\item \citet{lounici2011oracle}
\paragraph{Random matrix theory in machine learning.}
This work joins an emerging line of recent work that has found random matrix theory useful for studying modern machine learning techniques.
A prominent example is looking at interpolators in over-parametrized linear and logistic regression \cite{bartlett2020benign,hastie2019surprises,montanari2019generalization,liang2020just,liang2020precise}.
Interpolators can reproduce similar phenomena observed under neural networks because both models learn by ``interpolating'' the data using many parameters \cite{mei2019generalization}.
The precise rates achieved via random matrix theory can also explain the double-descent phenomenon \cite{belkin2019reconciling}.
Whereas these works deal with covariates drawn from a single distribution, our work instead tackles distribution shift given covariates sampled from two distributions.
